Reviving the Lost Art of \\Combinator Graph Reduction

### 2023-01-31 experiments/superg/doc/paper/paper.tex
Finish up preliminaries.

\section{Preliminaries}
\label{sec:prelim}
\subsection{Lambda Calculus}
For an accessible introduction to the Lambda Calculus, we can recommend~\cite{rojas_tutorial_2015}.
Most introductory resources, including~\cite{rojas_tutorial_2015}, present Lambda Calculus with named bindings.
Kiselyov's algorithms~\cite{kiselyov_lambda_2018} use De Bruijn indices~\cite{de_bruijn_lambda_1972} instead.

@@

@@

Any lambda calculus expression can be compiled using this algorithm.

TODO more!

>>>>

\section{Preliminaries}
\label{sec:prelim}
\subsection{Lambda Calculus}
For an accessible introduction to the Lambda Calculus, we can recommend~\cite{rojas_tutorial_2015}.

The Lambda Calculus comes in many flavors.
The most basic version has only function abstraction and application, and is untyped.
Kiselyov's description of his new algorithms~\cite{kiselyov_lambda_2018} use a typed variant.
Our description of the translation is not defined as a type system, hence the untyped Lambda Calculus suffices for our treatment.
We do, however, enrich the Lambda Calculus with a few primitives that are convenient in a practical implementation. They are:
\begin{itemize}
    \item Machine integers (signed 32-bit). These can be emulated with e.g. Church Encoding, but arithmetic over a native integer type is faster.
    \item Arithmetic operations. To manipulate machine integers, we introduce the usual $+$, $-$, $*$ and $/$ operations.
    \item Comparison operations. To compare machine integers, we also add $==$, $<$ etc. These operations return $1$ if true, and $0$ if false (machine integers do double duty as a boolean type).
\end{itemize}

Most introductory resources, including~\cite{rojas_tutorial_2015}, present Lambda Calculus with named bindings.
Kiselyov's algorithms~\cite{kiselyov_lambda_2018} use De Bruijn indices~\cite{de_bruijn_lambda_1972} instead.
As this is core to the algorithm, we adopt the same approach.
However, for example expressions, we continue to use the easier-to-read named bindings.

@@

Combinatory Logic is a notation designed to remove variables (named or otherwise) from Lambda Calculus expressions.
The core idea is to define a set of simple basic functions, or \emph{combinators}, that can be combined to pass variables to the proper subexpressions.
We only need three combinators to convert any Lambda Calculus expression:
\begin{itemize}
    \item $S \ f \ g \ x = f \ x \ (g \ x)$: Passes an argument to both subexpressions of an application.
    \item $K \ x \ y = x$: Drops the second argument.
    \item $I \ x = x$\footnote{$SKK = I$, so even this combinator is not strictly necessary, but all reasonable implementations do include $I$.}: The identity function that accepts an argument and returns it.
\end{itemize}

The traditional algorithm to compile Lambda Calculus expressions into Combinator Logic with these three combinators is Bracket Abstraction, which we present in the next section.

The set of combinators used in the translation may be extended to generate smaller programs.
Kiselyov's algorithms~\cite{kiselyov_lambda_2018} for example use the following additional combinators as special cases of $S$:

\begin{itemize}
    \item $B\ f\ g\ x = f\ (g\ x)$: Passes an argument to the \emph{right} subexpression.
    \item $C\ f\ g\ x = (f\ x)\ g$: Passes an argument to the \emph{left} subexpression.
\end{itemize}

Another addition, specifically for the $\linear$ algorithm, is the addition of \emph{bulk} combinators, variants of $S$, $B$ and $C$ that pass multiple arguments:

\begin{itemize}
    \item $S_n\ f\ g\ x_1\ ...\ x_n = (f\ x_1\ ...\ x_n)\ (g\ x_1\ ...\ x_n)$
    \item $B_n\ f\ g\ x_1\ ...\ x_n = f\ (g\ x_1\ ...\ x_n)$
    \item $C_n\ f\ g\ x_1\ ...\ x_n = (f\ x_1\ ...\ x_n)\ g$
\end{itemize}

@@

Any lambda calculus expression can be compiled using this algorithm.
As an example, we can compile the expression $\lambda x.\lambda y. x$, and obtain:

\begin{equation*}
    \begin{array}{l l}
        C\ [[\ \lambda x.\lambda y. x\ ]]        & \Rightarrow \\
        A\ x\ [[\ C\ [[\ \lambda y. x\ ]]\ ]]    & \Rightarrow \\
        A\ x\ [[\ A\ y\ [[\ C\ [[\ x\ ]]\ ]]\ ]] & \Rightarrow \\
        A\ x\ [[\ A\ y\ [[\ x\ ]]\ ]]            & \Rightarrow \\
        A\ x\ [[\ K\ x\ ]]                       & \Rightarrow \\
        S\ (A\ x\ [[\ K\ ]])\ (A\ x\ [[\ x\ ]])  & \Rightarrow \\
        S\ (K\ K)\ I                             & \Rightarrow \\
    \end{array}
\end{equation*}

The generated expression consists exclusively of combinators, as expected.
We can show that it produces the same result as the original lambda expression by applying it to two abstract variables $x$ and $y$:

\begin{equation*}
    \begin{array}{l l}
        S\ (K\ K)\ I\ x\ y   & \Rightarrow \\
        (K\ K\ x)\ (I\ x)\ y & \Rightarrow \\
        K\ x\ y              & \Rightarrow \\
        x                    &             \\
    \end{array}
\end{equation*}

While correct, the Bracket Abstraction algorithm is not optimal.
Looking at the derivation above, $S\ (K\ K)\ I$ is identical to the $K$ combinator.
In Section~\ref{sec:kiselyov}, we cover the better compilation algorithms proposed in~\cite{kiselyov_lambda_2018}.
